Inference using context-dependent models of mutation

### 2016-07-17 writeups/context-dependence.tex
added writeup of computing full likelihoods from abphylo

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\given}{\;\mid\;}

@@

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\given}{\;\mid\;}
\newcommand{\like}{\mathcal L}
\newcommand{\loglike}{\ell}
\newcommand{\alike}{\widetilde{\like}}
\newcommand{\aloglike}{\widetilde{\loglike}}

@@

%%%%%%
\section{Full likelihood}

\emph{The following is copied over from work on the abphylo project.}

The s5f model is defined as follows:
for each $(a,b,c,d,e,x) \subset \{\nA,\nC,\nG,\nT\}$,
it gives a (relative) mutation rate
\begin{align}
    \gamma(c,x;a,b,d,e) = \text{( mutation rate of $c \to x$ in context $abcde$ )} .
\end{align}
Given these rates, the only unknown parameter is the total amount of time
that the sequence has been mutated for.
Suppose that $X(t)$ is a sequence of length $n$ evolving under this model,
and let $G_n$ be the corresponding generator matrix,
a $4^n \times 4^n$ matrix.
(In the s5f model, the mutation rates are only specified in the middle of 5-mers,
so the two bases on each end never mutate.)
The transition probabilities are $e^{tG_n}$,
and by marginalization,
we can compute quantities like
the probability for observing the middle set of basis
given the flanking $\ell$ on each end:
\begin{align}
    p_{\ell,n-2\ell}(t;a,b)
    &=
    \P\{
        (X_{\ell+1}(t), X_{\ell+2}(t), \ldots, X_{n-\ell}(t))=(b_1,b_2,\ldots,b_{n-2\ell}) \\
    &\qquad \qquad
        \given (X_{1}(0), X_{2}(0), \ldots, X_{n}(0))=(a_1,a_2,\ldots,a_{n})
    \} .
\end{align}
If the size of the context, $\ell$, is big enough,
then this probability will not depend significantly if the sequence $a$ is in fact embedded in a longer sequence
(the error decreases exponentially with $\ell$).

The sequence $X(t)$ can be constructed by beginning with $X(0)$,
putting down a Poisson process of potential mutations of rate $\gamma = \max\{\gamma(c,x;a,b,d,e)\}$,
and thinning appropriately.
The process depends on the flanking 2bp,
so any other mutation within 2bp of a site affects the process at that site,
and the chance that this occurs depends on the flanking 4bp on either side.
Concretely, the number of flanking positions in each direction
that must be kept track of to decide what happens at a given site
grows by 1 at rate $\gamma$, and by 2 at rate $\gamma$;
so the amount of flanking sequence on each side affecting a given site at time $t$
is the sum of a Poisson($\gamma t$) and twice an independent Poisson($\gamma t$).
This has mean $3\gamma t$ and variance $5 \gamma t$;
with high probability it will not exceed, say,
$3\gamma t + 6 \sqrt{\gamma t}$;
if $\gamma t < 1$ this is less than 9.

How can we use this to compute the likelihood of a full sequence?
Actually, we know more.
For two positions $(i,s)$ and $(j,t)$ in sequence $\times$ time,
let $r(i,s;j,t) = |j-i|/|t-s|$ be the slope of the line between them,
and if $I$ and $J$ are subsets of sites, let $r(I,s;J,t) = \min\{r(i,s;j,t) : i\in I, j \in J\}$
be the minimum.
Then, we know that there is a speed of dependency propagation $r_*$
such that if $r(I,s;J,t) > r_*$,
then $X_I(s)$ and $X_J(t)$ are nearly independent.
Now, fix $t$ and for each site $i$ let $N_i$ denote its $r_*$-neighborhood,
i.e.,
\begin{align}
    N_i = \{ (j,s) : r(i,t; j,s) \le r_*, \; 0 \le s \le t \} .
\end{align}
These are depicted in figure~\ref{fig:neighborhoods}A.
We will write $i \sim j$ if $|i-j| \le 2 r_* t$,
or equivalently, if $N_i \cap N_j \neq \emptyset$,
and $i \nsim j$ otherwise.
We can think of $N_i$ as the parts of ancestral sequences in which a mutation might have affected the state at $(i,t)$.
so if $i \nsim j$, then $i$ and $j$ are (approximately) independent at time $t$, given the sequence at time 0.
Concretely,
we have a small $\epsilon$ such that for any $i \nsim j$,
\begin{align}
    &\big|
    \log( \P\{ X_i(t) = x_i \and X_j(t) = x_j \} ) \\
    &\qquad {} -
    \log( \P\{ X_i(t) = x_i \} \P\{ X_j(t) = x_j \} )
    ) \big|
    \le \epsilon .
\end{align}

\begin{figure}
    \begin{center}
        \includegraphics{context-neighborhoods-defns}

        \vspace{2em}

        \includegraphics{context-neighborhoods-widths}
    \end{center}
    \caption{
        Diagram of things in the text:
        (top) Dependency neighboorhods.
        (bottom) What we need to compute $\P\{X_i(t)=x \given X(0), X_{<i}(t)\}$:
        site $i$ depends on the blue squares in $N_i$;
        the sites to the left of this in $L(i)$ are those other sites in $X(t)$
        whose dependency neighborhoods overlap $N_i$;
        any further sites to the left of this are basically independent of $X_i(t)$;
        the red squares give the dependency neighborhood of $L(i)$,
        so we need to condition on the row of red and blue sites at the top.
        On the bottom, we have $2w+1$ sites; at the top we have $4w+1$;
        here $w=2$.
        \label{fig:neighborhoods}
    }
\end{figure}

What we'd like to know is the likelihood, i.e.,
\begin{align}
    \like(y|x)
    &:=
    \P\{ X(t) = y \given X(0) = x \} .
\end{align}
If we define $X_{<i}=(X_1, \ldots, X_{i-1})$, then
\begin{align}
    \like(y|x)
    &=
    \prod_{i=1}^n
    \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} .
\end{align}
Now, defining $L(i) = \{ j < i \st j \sim i \}$
to be the sites to the left of $i$ whose dependency neighborhoods overlap with $i$'s,
and $A(I) = \{ j \st |j-i| \le r_*t \text{ for some } i \in I\}$
to be the ancestral sites in the dependency neighborhood of $I$,
\begin{align}
&\big|
\log( \P\{ X_i(t) = y_i \given X(0) = x \and X_{<i}(t) = y_{<i} \} ) \\
& \qquad {} -
\log( \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} )
 \big| \\
&\qquad \qquad \le
    \epsilon ,
\end{align}
so that if we define
\begin{align}
    \alike(y|x)
    &=
    \prod_{i=1}^n
    \P\{ X_i(t) = y_i \given X_{A(L(i))}(0) = x_{A(L(i))} \and X_{L(i)}(t) = y_{L(i)} \} ,
\end{align}
then (with $\loglike = \log \like$),
\begin{align}
    \left| \aloglike(y|x) - \loglike(y|x) \right|
    \le n \epsilon .
\end{align}

This is good, because we can compute $\aloglike$:
let $w=r_*t$ be radius of a dependency neighborhood,
so that the width of $L(i)$ is $2w$,
and for a sequence $x$ define $x^{(m)}_i = (x_{i-m+1}, \ldots, x_i)$:
\begin{align}
    \alike(y|x)
    &=
    \prod_{i=1}^n
    \frac{
        p_{w,2w+1}(y^{(2w+1)}_i|x^{(4w+1)}_{i+w})
    }{
        p_{w,2w}(y^{(2w)}_{i-1}|x^{(4w)}_{i-1+w})
    }
\end{align}
This is the ratio of the likelihoods of all the $(w,2w+1)$ T-mers,
divided by the likelihoods of all the $(w,2w)$ T-mers.
